---
title: "iClusterPlus"
output: pdf_document
bibliography: iClusterPlus_bib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

This is a tutorial from the iClusterPlus package (see iClusterPlus::iManual() 
for details).

First we call the relevant libraries.

```{r libraries, echo=T, message=FALSE}

# The packages available in CRAN
required_packages <- c(
  "BiocManager",
  "cluster",
  "gplots",
  "lattice",
  "dplyr"
)

# The packages available from Bioconductor
bioconductor_packages <- c(
  "iClusterPlus",
  "GenomicRanges",
  "DNAcopy"
)

# Check if the packages are avilable - if not install them
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

for (b_pkg in bioconductor_packages) {
  if (!requireNamespace(b_pkg, quietly = TRUE)) {
    BiocManager::install(b_pkg)
  }
}

# Load the required libraries
library(iClusterPlus)
library(GenomicRanges)
library(DNAcopy)
library(cluster) # used for finding medoids in the adaptive model fit
library(gplots)
library(lattice)
library(dplyr)
```

This tutorial uses the glioblastoma data from the Cancer Genome Atlas (TGCA)
[@cancer2008comprehensive].

```{r gbm_data}
data(gbm)
dim(gbm.mut)

mut.rate <- apply(gbm.mut, 2, mean)
gbm.mut2 <- gbm.mut[, which(mut.rate > 0.02)]
gbm.mut2[1:10, 1:8]
```

For gene expression data, we recommend using the top variable genes for
integrative clustering analysis, which can be obtained by variance filtering.
For example, we use the top 1740 genes for our iCluster analysis.

```{r gbm_expression}
# the rows and columns corresponding to the samples and genes respectively
gbm.exp[1:3, 1:8]
```

It is a challenge to incorporate raw or normalized copy number data for
iCluster analysis considering the high dimensionality and spatial correlation
of the data. Based on our experience, we think it is more feasible to use the
segmentation results produced by the DNAcopy package.

```{r gbm_segment}
dim(gbm.seg)
gbm.seg[1:3, ]
```

We reduce the GBM copy number regions to 5K by removing the redundant regions
using function CNregions.

```{r copy_num}
data(variation.hg18.v10.nov.2010)

# gbm.cn is the segmentation results produced by DNAcopy
gbm.cn <- CNregions(
  seg = gbm.seg,
  epsilon = 0,
  adaptive = FALSE,
  rmCNV = TRUE,
  cnv = variation.hg18.v10.nov.2010[, 3:5],
  frac.overlap = 0.5,
  rmSmallseg = TRUE,
  nProbes = 5
)
```

Here seg is the DNAcopy segmentation output. In the first step, we define a set 
of non-redundant regions with parameter epsilon that denotes the maximum 
distance (Euclidean) between adjacent probes tolerated for defining a
non-redundant region. epsilon=0 is equivalent as taking the union of all unique 
break points across the n samples. Default epsilon=0.005. We then take the 
medoid signature as the representative copy number profile for that region (the 
medoid is similar to the centroid of a dataset, but restricted to a point within
the dataset). The degree of dimension reduction is proportional to the number of
samples included. We recommend setting an epsilon such that the reduced 
dimension is less than 10K. When sample size is large, an adaptive dimension 
reduction is more effective.

Instead of setting absolute threshold epsilon, setting adaptive=T will reduce 
dimension proportional to upper quantile of the
distances (default is False.) rmCNV=T remove germline CNV. When set to True, one
must supply a list of germline CNVs as cnv=cnv. The additional argument 
rmSmallseg removes small segments likely to be noise. Default is False. When set
of True, nProbes, the segment length threshold below which rmSmallseg will 
exclude should be specified.


```{r example_adaptive_thresholding}

gbm.cn.adapat <- CNregions(
  seg = gbm.seg,
  epsilon = 0,
  adaptive = TRUE,
  rmCNV = TRUE,
  cnv = variation.hg18.v10.nov.2010[, 3:5],
  frac.overlap = 0.5,
  rmSmallseg = TRUE,
  nProbes = 5
)
```

Sort gbm.cn to make sure all the samples are in the same order.

```{r check_dataset_order}
gbm.cn <- gbm.cn[order(rownames(gbm.cn)), ]

# check if all the samples are in the same order for the three data sets
all(rownames(gbm.cn) == rownames(gbm.exp))
all(rownames(gbm.cn) == rownames(gbm.mut2))
```

## Integrative clustering analysis
iClusterPlus fits a regularized latent variable model based clustering that 
generates an integrated cluster assignment based on joint inference across data 
types. Below is a one-liner to run iClusterPlus given the desired number of
eigen-features k (number of clusters is k+1) and the values of parameter set 
lambda (which determines sparsity). Normally, we have to optimize k and lambda
through model tuning which we discuss in the next section.

```{r iClusterPlus_call, eval = F}
n_clusters <- 3
n_iter <- 10
data_types <- c("binomial", "gaussian", "gaussian")
lambdas <- c(0.04, 0.61, 0.90)

fit.single <- iClusterPlus(
  dt1 = gbm.mut2,
  dt2 = gbm.cn,
  dt3 = gbm.exp,
  type = data_types,
  lambda = lambdas,
  K = n_clusters - 1,
  maxiter = n_iter
)
```

## Model tuning
Using parallel computing, tune.iClusterPlus samples a series of lambda values 
from the parameter space based on the Uniform design [@fang1994uniform] to 
search for the best model. The number of points to sample (`n.lambda`) depends 
on the number of data types and can take the following values: If we know the 
number of clusters in the samples, we may just choose the corresponding $k$ (the
number of latent variables) for the cluster analysis. We use the rule that the 
number of clusters equals $k+1$. If we don’t know the cluster number, we can 
test the $k$ from $1$ to $N$ (a reasonable number of clusters). In this example,
we let the $k$ from 1 to 5.

We have a number of options for our choice of `n.lambda` depending on the number
of data.types present.

|  Number of data types|  `n.lambda`|
|--:|--:|
|  1|  any|
|  2|  8, 13, 21, 34, 55, 89, 144, 233, 377, 610|
|  3|  35, 101, 135, 185, 266, 418, 597, 828, 1010|
|  4|  307, 562, 701, 1019, 2129, 3001, 4001, 5003, 6007|

Here for demonstration purpose, we specify `n.lambda`$=185$ sampling points (a 
modest value for this setting) which may result in some variability. As a 
result, either $k=2$ can be chosen as the best $k$ based on the percentage of 
explained variation. For general practice, set `n.lambda=NULL` to use the 
default value.

```{r model_tuning, eval = F}

# This is slow. Setting to not evaluate in Markdown.

# Set seed for consistent results.
set.seed(1)

# iterate over k
K <- 5
for (k in 1:K) {
  cv.fit <- tune.iClusterPlus(
    cpus = 3,
    dt1 = gbm.mut2,
    dt2 = gbm.cn,
    dt3 = gbm.exp,
    type = c("binomial", "gaussian", "gaussian"),
    K = k,
    n.lambda = 185,
    scale.lambda = c(1, 1, 1),
    maxiter = 20
  )
  save(cv.fit, file = paste("cv.fit.k", k, ".Rdata", sep = ""))
}
```

## Model selection

An illustration of model tuning. To select the vector $\lambda$ of penalty 
parameters iClusterPlus iterates over the number of latent variables defined by 
$k$. Then the Bayesian information criterion (BIC) is used to select the optimal
sparse model with regards to combinations of penalty parameters. To optimise 
$k$, the deviance ratio is the metric of choice. For a given model defined by 
parameters $\theta_{fit}$, the deviance ratio is defined by the saturated model 
(the model for which there is a unique parameter for every observation) defined 
by $\theta_{sat}$, and the null model (the intercept only model) defined by 
$\theta_{null}$:

$$Dev(\theta_{fit}) = \frac{l(\theta_{fit}) - l(\theta_{null})}{l(\theta_{sat}) - l(\theta_{null})}.$$

This ratio can be interpreted as the percentage of explained variation. Using a 
plot of the deviance ratio a function of $k$, the value of $k$ for which the 
deviance ratio levels off is selected.

```{r optimise_k, eval = F}
output <- alist()
files <- grep("cv.fit", dir())
for (i in 1:length(files)) {
  load(dir()[files[i]])
  output[[i]] <- cv.fit
}
nLambda <- nrow(output[[1]]$lambda)
nK <- length(output)
BIC <- getBIC(output)
devR <- getDevR(output)
```

From this one obtains the ID for the lambda vector at which the BIC is minimised.
Then we obtain the corresponding deviance ratio.

```{r min_bic_dev, eval = F}
minBICid <- apply(BIC, 2, which.min)
devRatMinBIC <- rep(NA, nK)
for (i in 1:K) {
  devRatMinBIC[i] <- devR[minBICid[i], i]
}
```

We plot the deviance ratio as a function of the number of clusters ($k+1$).

```{r plot_k_scree, eval = F}
plot(1:(K + 1),
  c(0, devRatMinBIC),
  type = "b",
  xlab = "Number of clusters (K+1)",
  ylab = "% Explained Variation"
)
```

From this an optimal value of $k=2$ (i.e. $n_{clusters}=3$) is selected.

```{r get_clusters, eval = F}
clusters <- getClusters(output)
rownames(clusters) <- rownames(gbm.exp)
colnames(clusters) <- paste("K=", 2:(length(output) + 1), sep = "")
# write.table(clusters, file="clusterMembership.txt",sep='\t',quote=F)
k <- 2
best.cluster <- clusters[, k]
best.fit <- output[[k]]$fit[[which.min(BIC[, k])]]
```

For noisy data a curve as shown above is unlikely - the choice of $k$ will 
probably be a lot less clear. In this case it is recommended that one creates
heat maps of the datasets and decide the optimal number of clusters based upon 
the features' pattern.


## References
