---
title: "MATLAB MDI on simple data"
author: "Stephen Coleman"
date: "11/02/2020"
output: pdf_document
---

```{r setup, include=FALSE}
library(pheatmap)
library(mdiHelpR)
library(tibble)
library(magrittr)

col_pal <- colorRampPalette(c("white", "#146EB4"))(100)
col_pal_expr <- colorRampPalette(c("#146EB4", "white", "#FF9900"))(100)
my_breaks <- c(
  seq(0, 1, length.out = ceiling(length(col_pal)) + 1)
)
expr_breaks <- defineBreaks(col_pal_expr)
```

## Data
The data generated is derived from two multivariate distirbutions defined by their means:
\[
\mu_1 = \begin{pmatrix}
1 \\ 1 \\1 
\end{pmatrix}, \mu_2 = \begin{pmatrix}
-1 \\ -1 \\-1 
\end{pmatrix}.
\]
Both distributions use $\mathbf{I}$ for the covariance matrix.

## MDI

MATLAB MDI is run on two datasets for 50,000 iterations, thinning by a factor of 50. One can see that the labels indicating which distribution the data originates from aligns (almost) perfectly with the clustering captured in the PSM.

```{r read_in_data}
# Read in the tibble containing the MDI analysis
my_tib <- readRDS("./compare_tibble.rds")

# The original data
my_data <- my_tib$expression_data[[1]]

# Briefly summarise
summary(my_data)

# The PSM
my_psm <- my_tib$similarity_matrix[[1]]

# The labels for which subpopulation each person actually belongs to
labels <- as.factor(c(
  rep(1, 0.5 * nrow(my_data)),
  rep(2, 0.5 * nrow(my_data))
))

# These labels in an annotation dataframe for the pheatmap
annotation_row <- data.frame(Labels = labels) %>%
  set_rownames(row.names(my_data))

# Visualise the PSM annotated by true labelling
pheatmap(my_psm,
  annotation_row = annotation_row,
  color = col_pal,
  breaks = my_breaks,
  main = "PSM annotated by true labelling"
)
```

Comparing this PSM directly with the expression data:

```{r compare_psm_data}

# Extract the order from the pheatmap
row_order <- hclust(dist(my_psm))$order

# Find the column order for the expression data
col_order <- hclust(dist(t(my_data)))$order

# Re order the matrices to have a common row order
my_psm <- my_psm[row_order, row_order]
my_data <- my_data[row_order, col_order]

colnames(my_psm) <- NULL
colnames(my_data) <- NULL

# Save the heatmaps of these to the same grid
compareHeatmapPSMandData(my_psm, my_data,
  save_name = NULL,
  main = "Comparison of PSM and data",
  col_pal_sim = col_pal,
  col_pal_expr = col_pal_expr,
  expr_breaks = expr_breaks,
  sim_breaks = my_breaks
)
```
