---
title: "Weekly update 04/12/2019-11/12/2019"
author: "Stephen Coleman"
date: "10/12/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Consensus clustering

I have discovered that the command line version of MDI does not handle sparse categorical data well. I am trying to understand the priors he uses (the most likely source of the problem according to Paul), but a lack of comments apart from such cases as "TODO: add an explicit ``sample from prior" call in here" are making it an uphill battle.

### Comparison command line to MATLAB

```{r load_data, echo = FALSE, results='hide'}
library(tibble)
library(mdiHelpR)

# comparePSMs <- function(sim_1, sim_2,
#                         curr_title = "compare_psms",
#                         col_pal = c("#FFFFFF", grDevices::colorRampPalette(RColorBrewer::brewer.pal(n = 7, name = "Blues"))(100)),
#                         breaks = defineBreaks(col_pal, lb = 0, ub = 1)){
#
#   # Extract the order from the pheatmap
#   row_order <- hclust(dist(sim_1))$order
#
#   # Re order the matrices to have a common row order
#   sim_1 <- sim_1[row_order, row_order]
#   sim_2 <- sim_2[row_order, row_order]
#
#   colnames(sim_1) <- NULL
#   colnames(sim_2) <- NULL
#
#   # Save the heatmaps of these to the same grid
#   compareHeatmapPSMandData(sim_1, sim_2,
#                            save_name = NA,
#                            main = curr_title,
#                            col_pal_sim = col_pal,
#                            col_pal_expr = col_pal,
#                            expr_breaks = breaks,
#                            sim_breaks = breaks
#   )
# }

mdi_mason <- readRDS("./Data/cmd_line_358500.rds")
mdi_matlab <- readRDS("./Data/matlab_8080.rds")
consensus_mason <- readRDS("./Data/cmd_line_consensus_clustering.rds")
consensus_matlab_100 <- readRDS("./Data/matlab_consensus_clustering_100.rds")
consensus_matlab_500 <- readRDS("./Data/matlab_consensus_clustering_500.rds")

col_pal <- c("#FFFFFF", grDevices::colorRampPalette(RColorBrewer::brewer.pal(n = 7, name = "Blues"))(100))
breaks <- defineBreaks(col_pal, lb = 0, ub = 1)
```

```{r combine_pheatmaps, echo = FALSE}

dataset_1 <- mdi_mason$dataset[[1]]
dataset_2 <- mdi_mason$dataset[[2]]
dataset_3 <- mdi_mason$dataset[[3]]

n_datasets <- length(mdi_mason$dataset)
ph_list <- vector("list", 6 * n_datasets)

n_ph <- 8
for (i in 1:n_datasets) {
  dataset <- mdi_mason$dataset[[i]]

  mdi_mason_sim <- mdi_mason$similarity_matrix[mdi_mason$dataset == dataset][[1]]
  mdi_matlab_sim <- mdi_matlab$similarity_matrix[mdi_matlab$dataset == dataset][[1]]
  consensus_mason_sim <- consensus_mason$similarity_matrix[consensus_mason$dataset == dataset][[1]]
  consensus_matlab_100_sim <- consensus_matlab_100$similarity_matrix[consensus_matlab_100$dataset == dataset][[1]]
  consensus_matlab_500_sim <- consensus_matlab_500$similarity_matrix[consensus_matlab_500$dataset == dataset][[1]]

  mdi_mason_expr <- mdi_mason$expression_data[mdi_mason$dataset == dataset][[1]]
  mdi_matlab_expr <- mdi_matlab$expression_data[mdi_matlab$dataset == dataset][[1]]
  consensus_mason_expr <- consensus_mason$expression_data[consensus_mason$dataset == dataset][[1]]
  consensus_matlab_100_expr <- consensus_matlab_100$expression_data[consensus_matlab_100$dataset == dataset][[1]]
  consensus_matlab_500_expr <- consensus_matlab_500$expression_data[consensus_matlab_500$dataset == dataset][[1]]

  ph_list[[n_ph * (i - 1) + 1]] <- compareMatrices(mdi_matlab_sim,
    mdi_mason_sim,
    curr_title = paste0(dataset, ": MATLAB vs CMD Line"),
    col_pal = col_pal,
    col_names = F
  )

  ph_list[[n_ph * (i - 1) + 2]] <- compareMatrices(mdi_mason_sim,
    mdi_matlab_sim,
    curr_title = paste0(dataset, ": CMD Line vs MATLAB"),
    col_pal = col_pal,
    col_names = F
  )

  ph_list[[n_ph * (i - 1) + 3]] <- compareMatrices(mdi_matlab_sim,
    consensus_mason_sim,
    curr_title = paste0(dataset, ": MATLAB vs CMD Line consensus"),
    col_pal = col_pal,
    col_names = F
  )

  ph_list[[n_ph * (i - 1) + 4]] <- compareMatrices(mdi_matlab_sim,
    consensus_matlab_100_sim,
    curr_title = paste0(dataset, ": MATLAB vs MATLAB consensus (100)"),
    col_pal = col_pal,
    col_names = F
  )

  ph_list[[n_ph * (i - 1) + 5]] <- compareMatrices(consensus_matlab_100_sim,
    consensus_mason_sim,
    curr_title = paste0(dataset, ": MATLAB consensus (100) vs CMD Line consensus"),
    col_pal = col_pal,
    col_names = F
  )

  ph_list[[n_ph * (i - 1) + 6]] <- compareMatrices(consensus_mason_sim,
    consensus_matlab_100_sim,
    curr_title = paste0(dataset, ": CMD Line consensus vs MATLAB consensus (100)"),
    col_pal = col_pal,
    col_names = F
  )

  ph_list[[n_ph * (i - 1) + 7]] <- compareMatrices(mdi_matlab_sim,
    consensus_matlab_500_sim,
    curr_title = paste0(dataset, ": MATLAB MDI vs MATLAB consensus (500)"),
    col_pal = col_pal,
    col_names = F
  )

  ph_list[[n_ph * (i - 1) + 8]] <- compareMatrices(consensus_matlab_100_sim,
    consensus_matlab_500_sim,
    curr_title = paste0(dataset, ": MATLAB consensus (100) vs MATLAB consensus (500)"),
    col_pal = col_pal,
    col_names = F
  )
}
```

Let's compare the posterior similarity matrices (PSMs) produced by the different methods. Our methods are:

1. Command line MDI (358,500);
2. Command line consensus clustering (1,000 seeds for 500 iterations);
3. MATLAB MDI (8,080 iterations);
4. MATLAB consensus clustering (1,000 seeds for 100 iterations); and
5. MATLAB consensus clustering (1,000 seeds for 500 iterations).

The number of iterations is an odd number for the long chains due to constraints on the HPC.

In each case the MATLAB MDI is being considered our gold standard. This might not be fair as there's no strong reason to think this chain has converged, but it's similar to results to the original MDI paper, so for now it's a good heuristic for checking for sensible results.

We have three datasets, the time course data (which seems most consistent across methods), which is run using the Gaussian Process setting, and then two sparse categorical datasets, the Chip-chip transcription factor data from the Harbison paper and the protein-protein interaction data (Harbison and PPI resepective). Let's look at the comparison of the timecourse data for the MATLAB MDI and Command line MDI: 

```{r mdi_comp_timecourse, echo=FALSE}
ph_list[[1]]
```

Comparing the MATLAB to both forms of consensus clustering:

```{r consensus_comp_timecourse, echo=FALSE}
ph_list[[3]]
ph_list[[4]]
ph_list[[7]]
ph_list[[8]]
```

This is not too bad. There's slightly different stories but quite a lot of agreement too. The real difference is within the categorical datasets:

```{r mdi_comp_categorical, echo = FALSE}
ph_list[[n_ph + 1]]
ph_list[[n_ph + 3]]
ph_list[[n_ph + 4]]
ph_list[[n_ph + 7]]
ph_list[[n_ph + 8]]

ph_list[[2 * n_ph + 1]]
ph_list[[2 * n_ph + 3]]
ph_list[[2 * n_ph + 4]]
ph_list[[2 * n_ph + 7]]
ph_list[[2 * n_ph + 8]]
```

We can see that the consensus clustering works pretty well even with only 100 iterations, but really 500 is a fair bit better.